\documentclass[11pt,a4paper]{article}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{float}
\usepackage{hyperref}
\usepackage{xcolor}

\title{Lightweight Hybrid Quantum-Classical Networks for Biomedical Tabular Binary Classification:\\
A Knowledge Distillation and Quantum Layer Placement Study}

\author{Xiaoming Liu\\
Panda Intelligence Software LTD, United Kingdom\\
\texttt{i@pandacat.ai}}

\date{}

\begin{document}
\maketitle

% -------------------------------------------------------
\begin{abstract}
Hybrid quantum-classical architectures offer a pragmatic path toward deploying quantum components
under near-term resource constraints. We propose a lightweight framework that combines a classical
teacher model, a compact student network, and a small parameterized quantum circuit (PQC) for
biomedical tabular binary classification under low-resource simulation. To make simulation
feasible on classical hardware, we apply PCA-based dimensionality reduction before quantum
encoding, reducing the required qubit count to 4. We train student models using a weighted
combination of hard-label and soft-target knowledge distillation (KD) loss, and systematically
compare classical and hybrid students with and without KD. We further ablate the effect of
quantum layer insertion position (front / middle / tail), input feature dimension, and distillation
hyperparameters on predictive performance and computational cost. Evaluated on the Breast Cancer
Wisconsin Diagnostic dataset with 5-fold cross-validation, our results show that [PLACEHOLDER:
key finding]. We report AUC, F1, MCC, training time, qubit count, and circuit depth to provide
a complete performance-cost trade-off picture. Our goal is not to claim quantum advantage, but
to identify practical design guidelines for hybrid quantum-classical learning in realistic
biomedical settings.
\end{abstract}

% -------------------------------------------------------
\section{Introduction}
\label{sec:intro}

Quantum machine learning (QML) has attracted growing interest as a bridge between near-term
quantum devices and practical machine learning tasks. However, under current NISQ constraints,
large quantum circuits remain difficult to train and expensive to simulate~[CITE]. As a result,
hybrid quantum-classical (HQC) architectures have emerged as a more realistic direction, combining
classical modules for feature compression with small parameterized quantum circuits for
learnable transformation~[CITE].

Biomedical tabular classification is a particularly suitable testbed for HQC models. Such tasks
are often low-dimensional after feature selection, clinically interpretable, and benefit from
compact models with competitive performance. Recent work has demonstrated the feasibility of
quantum-enhanced biomarker classification using QSVC and QNN with PCA-reduced inputs~[CITE],
suggesting that the combination of classical dimensionality reduction and quantum learning
components is a viable direction.

Meanwhile, knowledge distillation (KD) has proven effective for training compact student models
by leveraging soft supervisory signals from a larger teacher~[CITE]. Recent work has begun to
explore KD as a bridge between classical and quantum neural networks~[CITE], showing that
classical-to-quantum knowledge transfer can improve student training and reduce the effective
model complexity required.

Despite these advances, two questions remain underexplored:
\begin{enumerate}
\item Does KD consistently improve hybrid quantum-classical student models on biomedical tabular tasks?
\item Where should the quantum layer be placed within the student network for the best
      performance-to-cost trade-off?
\end{enumerate}

To address these questions, we make the following contributions:
\begin{enumerate}
\item We propose a lightweight HQC student framework with three quantum layer insertion positions
      (front, middle, tail) and evaluate their effect on AUC, F1, MCC, and training cost.
\item We introduce classical-to-hybrid KD into this framework and demonstrate its effect on
      model convergence and predictive performance.
\item We conduct systematic ablations on PCA dimension, distillation temperature, and
      insertion position under a consistent 5-fold cross-validation protocol.
\item We provide complete performance-cost reporting including qubit count, circuit depth,
      training time, and inference time to support reproducible comparison.
\end{enumerate}

% -------------------------------------------------------
\section{Related Work}
\label{sec:rw}

\subsection{Hybrid Quantum-Classical Classification}
Hybrid quantum-classical networks combine classical feature extraction with small parameterized
quantum circuits (PQC) for classification~[CITE]. Several works have studied this architecture
for image and text data~[CITE], but biomedical tabular data has received comparatively less
attention. Alvi et al.~[CITE] demonstrated that QSVC and QNN with PCA-reduced inputs can
achieve competitive performance on biomarker datasets, and that feature map choice and
class balance significantly affect quantum model outcomes.

\subsection{Knowledge Distillation for Quantum Models}
Knowledge distillation (KD) was originally proposed to transfer knowledge from a large teacher
to a compact student using soft logit targets~[CITE]. Recent work has extended this to
classical-to-quantum settings, showing that quantum students can benefit from classical teacher
supervision to improve training stability and generalization~[CITE]. QuantumMedKD~[CITE]
further applied hybrid KD to a medical imaging task, motivating our extension to tabular data.

\subsection{Biomedical Tabular Classification}
Tabular biomedical data (clinical assays, omics, biomarkers) presents challenges including
class imbalance, small sample sizes, and high feature dimensionality~[CITE]. PCA-based
feature compression has been shown effective for reducing dimensionality while preserving
classification-relevant variance, and is especially suitable for quantum model inputs that
require bounded feature dimensions~[CITE].

% -------------------------------------------------------
\section{Method}
\label{sec:method}

\subsection{Problem Setup}
Given a biomedical tabular dataset $\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^N$ with binary
labels $y_i \in \{0, 1\}$ and continuous feature vectors $\mathbf{x}_i \in \mathbb{R}^d$, we
aim to learn a compact classifier that achieves competitive AUC and F1 while minimizing
computational cost (training time, qubit count, circuit depth).

\subsection{Preprocessing}
Within each training fold, we apply:
\begin{enumerate}
\item Standard scaling (zero mean, unit variance).
\item PCA to reduce input dimensionality to $d' \in \{4, 6, 8\}$.
\item Optional class balancing (class weights or SMOTE).
\end{enumerate}
To prevent data leakage, all preprocessing is fit on the training split only and applied
to the validation split.

\subsection{Teacher Model}
The teacher is a classical MLP with hidden dimensions $[64, 32]$, dropout 0.2, and sigmoid
output. It is trained with binary cross-entropy loss and provides soft logits
$\mathbf{z}^T \in \mathbb{R}^N$ for knowledge distillation.

\subsection{Student Models}
We consider two student architectures:

\textbf{Classical student (S-C):} A lightweight MLP with hidden dimensions $[16, 8]$.

\textbf{Hybrid quantum-classical student (S-H):} A three-segment architecture:
\begin{enumerate}
\item Classical front-end: projects input to a compact representation.
\item Quantum layer: a PQC with $n_q = 4$ qubits and $L = 2$ variational blocks,
      using angle encoding (RY) and RY/RZ rotations with CNOT entanglement.
\item Classical head: maps quantum expectation values to binary logits.
\end{enumerate}
We evaluate three insertion positions:
\begin{itemize}
\item \textbf{Front}: quantum layer receives PCA-reduced input directly.
\item \textbf{Middle}: classical projection first, then quantum layer.
\item \textbf{Tail}: classical projection + nonlinear layer, then quantum layer.
\end{itemize}

\subsection{Quantum Layer Design}
The PQC uses angle encoding to map $\mathbf{x} \in \mathbb{R}^{n_q}$ into qubit rotations:
\begin{equation}
|\psi_0\rangle = \bigotimes_{i=1}^{n_q} R_Y(x_i)|0\rangle
\end{equation}
Followed by $L$ variational blocks:
\begin{equation}
U_l(\boldsymbol{\theta}_l) = \left(\prod_{i=1}^{n_q-1} \text{CNOT}_{i,i+1}\right)
\cdot \left(\bigotimes_{i=1}^{n_q} R_Z(\theta_{l,i,1}) R_Y(\theta_{l,i,0})\right)
\end{equation}
Measurements return $n_q$ expectation values $\langle Z_i \rangle$ that are passed to the
classical head.

\subsection{Knowledge Distillation Loss}
The student is trained with a combined loss:
\begin{equation}
\mathcal{L} = (1-\lambda)\,\mathcal{L}_{\text{hard}} + \lambda T^2\,\mathcal{L}_{\text{KD}}
\label{eq:kd_loss}
\end{equation}
where $\mathcal{L}_{\text{hard}} = \text{BCE}(\hat{y}, y)$ is the hard-label binary
cross-entropy, $\mathcal{L}_{\text{KD}} = \text{KL}(\sigma(\mathbf{z}^S/T) \| \sigma(\mathbf{z}^T/T))$
is the soft-target KL divergence at temperature $T$, and $\lambda \in [0,1]$ controls the
distillation weight. When $\lambda = 0$, the student is trained without distillation.

\subsection{Evaluation Protocol}
All models are evaluated with stratified 5-fold cross-validation. We report mean $\pm$ std
over 5 folds for: AUC, F1, Accuracy, MCC, Precision, Recall, training time (s), inference
time (ms), qubit count, and circuit depth.

% -------------------------------------------------------
\section{Experiments}
\label{sec:exp}

\subsection{Dataset}
The Breast Cancer Wisconsin (Diagnostic) dataset~[CITE] contains 569 samples, 30 continuous
features, and binary labels (Malignant/Benign, 212/357 split). There are no missing values.
After PCA with $d' = 4$ components, the explained variance ratio is [PLACEHOLDER]\%.

\begin{table}[t]
\caption{Dataset statistics.}
\label{tab:data}
\centering\small
\begin{tabular}{lc}
\toprule
Property & Value \\
\midrule
Samples & 569 \\
Features (raw) & 30 \\
Features (after PCA) & 4 / 6 / 8 \\
Positive class (M) & 212 (37.3\%) \\
Negative class (B) & 357 (62.7\%) \\
Missing values & 0 \\
Task & Binary classification \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Model Configurations}

\begin{table}[t]
\caption{Experimental groups.}
\label{tab:groups}
\centering\small
\begin{tabular}{llcc}
\toprule
ID & Model & KD & Quantum Layer \\
\midrule
G1 & Teacher-Classic & -- & No \\
G2 & Student-Classic & No & No \\
G3 & Student-Classic-KD & Yes & No \\
G4 & Student-Hybrid-NoKD & No & Yes \\
G5 & Student-Hybrid-KD & Yes & Yes \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Main Results}

\begin{table}[t]
\caption{Main results on WDBC (5-fold CV, PCA $d'=4$, $T=2$, $\lambda=0.5$).}
\label{tab:main}
\centering\small
\begin{tabular}{lccccc}
\toprule
Model & AUC & F1 & MCC & Train Time (s) & Qubits \\
\midrule
Teacher-Classic & -- & -- & -- & -- & 0 \\
Student-Classic & -- & -- & -- & -- & 0 \\
Student-Classic-KD & -- & -- & -- & -- & 0 \\
Student-Hybrid-NoKD & -- & -- & -- & -- & 4 \\
Student-Hybrid-KD & -- & -- & -- & -- & 4 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Ablation: Quantum Layer Position}

\begin{table}[t]
\caption{Ablation: quantum layer insertion position (Student-Hybrid-KD, PCA $d'=4$).}
\label{tab:ablation_pos}
\centering\small
\begin{tabular}{lccccc}
\toprule
Position & AUC & F1 & MCC & Train Time (s) & Circuit Depth \\
\midrule
Front  & -- & -- & -- & -- & -- \\
Middle & -- & -- & -- & -- & -- \\
Tail   & -- & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Ablation: PCA Dimension and Distillation Hyperparameters}

\begin{table}[t]
\caption{Ablation: PCA dimension and distillation temperature (Student-Hybrid-KD, middle position).}
\label{tab:ablation_pca}
\centering\small
\begin{tabular}{lcccc}
\toprule
Setting & Value & AUC & F1 & Train Time (s) \\
\midrule
PCA dim & 4  & -- & -- & -- \\
PCA dim & 6  & -- & -- & -- \\
PCA dim & 8  & -- & -- & -- \\
$T$     & 2  & -- & -- & -- \\
$T$     & 4  & -- & -- & -- \\
$\lambda$ & 0.3 & -- & -- & -- \\
$\lambda$ & 0.5 & -- & -- & -- \\
$\lambda$ & 0.7 & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

% -------------------------------------------------------
\section{Discussion}
\label{sec:disc}

\subsection{Does KD help hybrid students?}
[PLACEHOLDER: Discuss G4 vs G5. Does KD improve AUC / F1 / MCC? Does it speed up convergence?]

\subsection{Which quantum layer position is best?}
[PLACEHOLDER: Compare front / middle / tail from Table~\ref{tab:ablation_pos}.
Discuss performance vs circuit depth trade-off.]

\subsection{Is the hybrid student worth the cost?}
[PLACEHOLDER: Compare G2 vs G4/G5 on performance-cost scatter.
Under what conditions does the quantum layer add value?]

\subsection{Limitations}
This study evaluates a single dataset under ideal statevector simulation using PennyLane.
Results may differ under shot noise, hardware noise, or on datasets with different statistical
properties. We do not claim quantum advantage; our goal is to identify practical design
guidelines within a reproducible simulation framework.

% -------------------------------------------------------
\section{Conclusion}
\label{sec:conc}

We presented a lightweight hybrid quantum-classical framework for biomedical tabular binary
classification, combining PCA-based dimensionality reduction, knowledge distillation, and
small parameterized quantum circuits. Through systematic ablation of quantum layer insertion
position, input dimension, and distillation hyperparameters, we identified the key factors
that govern the performance-cost trade-off of hybrid student models in a low-resource
simulation setting.

Future work includes: (1) evaluation on additional biomedical tabular datasets;
(2) extension to multi-class classification; (3) noise-aware training under hardware error
models; and (4) exploration of alternative quantum encodings and ansatz designs.

% -------------------------------------------------------
\section*{Acknowledgment}
We thank Panda Intelligence Software LTD for supporting this research.

\begin{thebibliography}{99}
\bibitem{wdbc} W.~Wolberg, O.~Mangasarian, N.~Street, and W.~Street, ``Breast Cancer Wisconsin (Diagnostic),'' UCI ML Repository, 1995.
\bibitem{kd} G.~Hinton, O.~Vinyals, and J.~Dean, ``Distilling the Knowledge in a Neural Network,'' arXiv:1503.02531, 2015.
\bibitem{biomarker_qml} [CITE: biomarker quantum classification paper]
\bibitem{c2q_kd} [CITE: classical to quantum KD paper]
\bibitem{quantummedkd} [CITE: QuantumMedKD paper]
\bibitem{pennylane} V.~Bergholm et al., ``PennyLane: Automatic differentiation of hybrid quantum-classical computations,'' arXiv:1811.04968, 2018.
\end{thebibliography}

\end{document}
